\section{Related Work}
\label{sec:related}

\paragraph{Output filters / content moderators.}
Llama Guard \citep{llamaguard} and WildGuard \citep{wildguard} are LLM-based
classifiers that label a single prompt or response as safe / unsafe. They are
effective content moderators but are designed for terminal text, not for gating
the next tool call in a multi-step agent loop. They have no notion of
trajectory state and cannot intervene mid-run; TraceGuard's unit of decision is
a proposed \emph{action} conditioned on the prior trace, and its intervention
is \emph{online}.

\paragraph{Post-hoc trajectory judges.}
R-Judge \citep{rjudge} evaluates whether an LLM can judge the safety of a
\emph{completed} agent trajectory, and benchmarks frontier models as graded
judges. This is valuable for offline auditing and for measuring judgment
ability, but a post-hoc judge cannot prevent the harmful action --- by the time
it scores the transcript, the action has executed. TraceGuard targets the
opposite regime: a small model that decides \emph{before} each step, in time to
block it.

\paragraph{Policy-rule agent shields.}
ShieldAgent \citep{shieldagent} enforces safety via explicit, often
LLM-assembled, policy rules over agent actions. Rule-based shields are precise
and auditable when a relevant rule exists, but they require the rule to be
written and tend to miss attacks outside the encoded policy. TraceGuard instead
learns the verdict distribution from labeled trajectories, so novel attack
phrasings that match learned patterns can be flagged without an explicit rule;
the trade-off is the usual one (learned coverage vs. auditable guarantees).

\paragraph{Program-analysis / symbolic shields.}
AgentArmor \citep{agentarmor} applies static / symbolic analysis over tool
calls and data flow (e.g. tracking tainted inputs to sensitive sinks). This is
complementary to a distributional monitor: symbolic taint tracking is strong
where tool semantics are well-typed, whereas TraceGuard's learned head can flag
anomalous \emph{patterns} even when individual calls look type-valid.

\paragraph{Trajectory anomaly detection.}
TrajAD-style \citep{trajad} approaches detect anomalous agent trajectories,
typically framed around \emph{quality} anomalies and within a single agent
framework. TraceGuard differs along three axes: (i) the target is
\emph{security} attacks, not quality drift; (ii) training data is
self-generated and folded into a \emph{framework-agnostic} canonical schema, so
one model spans frameworks; and (iii) intervention is \emph{online} via a
per-step verdict head, not a post-hoc classifier.

\paragraph{Agent attack benchmarks and datasets.}
AgentDojo \citep{agentdojo} provides a dynamic environment for benchmarking
prompt-injection attacks and defenses on tool-using agents, and ATBench
\citep{atbench} provides labeled agent trajectories with a three-tier taxonomy
(risk source $\times$ failure mode $\times$ harm category). We adopt ATBench's
taxonomy directly as our label space (Section~\ref{sec:data}) so that joint
evaluation is straightforward, and we draw on AgentDojo-style injection
patterns when synthesizing adversarial traces. Our model backbone is Qwen3.5-2B
\citep{qwen3}.

\paragraph{Summary of positioning.}
Relative to prior work, TraceGuard occupies a specific cell: a \emph{small},
\emph{learned}, \emph{online}, \emph{framework-agnostic} monitor evaluated under
a \emph{deployment} operating point (PR@FSR). Table~\ref{tab:positioning}
summarizes the comparison. % TODO: numbers in this table are qualitative; quantitative head-to-head on a shared benchmark is future work.

\begin{table}[t]
  \centering
  \small
  \begin{tabular}{lllll}
    \toprule
    System & Paradigm & Granularity & Online & Needs explicit rule \\
    \midrule
    Llama Guard / WildGuard & output filter      & terminal text & no  & no \\
    R-Judge (judges)        & post-hoc judge     & full trace    & no  & no \\
    ShieldAgent             & policy rules       & per action    & yes & yes \\
    AgentArmor              & symbolic analysis  & per action    & yes & partial \\
    TrajAD                  & learned classifier & trace         & no$^{\dagger}$ & no \\
    \textbf{TraceGuard}     & \textbf{learned verdict head} & \textbf{per step} & \textbf{yes} & \textbf{no} \\
    \bottomrule
  \end{tabular}
  \caption{Qualitative positioning. $^{\dagger}$Typically check-and-act /
  post-hoc rather than mid-trajectory veto. % TODO verify per-system claims against latest papers.
  }
  \label{tab:positioning}
\end{table}
